---
title: "01 EDA"
author: 
  - Patrick Reza Schnurbusch
  - Mike Mcowan
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: "Our technical analysis regarding the treaties"
output: 
  html_document:
    theme: flatly
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import_libraries, include=FALSE}
install.packages("wordcloud2")
install.packages("extrafont")
install.packages("igraph")
install.packages("plotly")
install.packages("quanteda")
install.packages("quanteda.textstats")
install.packages("topicmodels")
install.packages("wordcloud")
        
library(tm)
library(kernlab)
library(RPMG)
library(tis)
library(pastecs)
library(strucchange)

library(ggplot2)

library(wordcloud2) # for word clouds 
library(extrafont) # for system fonts 
library(igraph) # for network analysis 


library(plotly) # for graphing 
library(dplyr) # for data manipulation 
library(tidyr) # for pivot_wider() 


library(quanteda) # for NLP  
library(quanteda.textstats) # for readability score
library(topicmodels) # for topic modelling
library(wordcloud) # for word clouds 
```


```{r}
# list for plotly margins 
margin_list <- list(l = 100, r = 100, b = 100, t = 100, pad = 4)


# tile font 
title_font_list <- list(
  family = "Raleway", 
  size = 18, 
  color = "black"
)

font_list <- list(
  family = "Raleway", 
  size = 12, 
  color = "black"
)
```

# Background

This is a exploratory data analysis examining Prof. Spirling's findings from his 2012 Paper [U.S. Treaty Making with American Indians: Institutional Change and Relative Power, 1784–1911.](https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1540-5907.2011.00558.x)

Our goal is to first replicate his methodology, study his analysis, and then extend it to the Canadian context. After this is complete we will be positioned to begin performing comparative analysis between the two data sets.

The data we will be examining are the treaties (and manifests) for US Treaty-making with American Indians (AJPS), can be found in our `00_data` folder inside `original_treaties_usa`.

Note that they are divided into 4 groups as per the original paper describes:

1.  *Valid and Operable (VCUT)*

2.  *Ratified agreements (ACUT)*

3.  *Rejected (RCUT)*

4.  *Unratified (UCUT).*

A description of this data is found in `UniverseCases.csv`

The relevant data for the Indian Wars (as per Section 5.1) is found in `IndianWars.csv`

Below we will import the various datasets into our session, code folding is hidden by default by clicking on `code` or `show` will flip out the code for closer inspection.

We will be repeating the same methods in this analysis across all categories. 

TF-IDF (Term Frequency-Inverse Document Frequency), is a scoring method to determine how important a word is to the total document. We use this to weigh words that we find with higher significance based on how frequently they appear across our entire document corpus. We can use this to adjust for words that appear more frequently in general. 

# Valid and Operable (VCUT) {.tabset .tabset-fade .tabset-pills}

```{r}
document_cat <- "Valid & Operable"
```


365 documents compose our valid/operable category.

```{r}
# Valid and Operable (VCUT)
valid_operable <- tm::Corpus(
  x = tm::DirSource("./00_data/original_treaties_USA/justdocsVCUT/"),
  readerControl = list(
    reader = readPlain,
    language = "en",
    load = FALSE
  )
)

valid_operable
```

```{r}
# First, preprocess the data: remove common stop words, punctuation, and convert to lowercase
corpus <- valid_operable
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeWords, stopwords("english"))

# Create a term-document matrix
tdm <- TermDocumentMatrix(corpus)

# Convert the matrix to a data frame suitable for plotting
m <- as.matrix(tdm)
word_freqs <- sort(rowSums(m), decreasing = TRUE)
df <- data.frame(word = names(word_freqs), freq = word_freqs)
```


## Text Statistics 

```{r}
top_words <- df %>% dplyr::arrange(desc(freq)) %>% dplyr::slice_head(n = 50)

# order bars 
top_words$word <- factor(top_words$word,
                                      levels = unique(top_words$word)[order(top_words$freq, decreasing = F)])

plot_ly(data = top_words, y = ~word, x = ~freq, 
        type = 'bar', 
        hoverinfo = "text", 
        textposition = "auto", 
        hovertext = ~paste0("<b>",word,"</b>","<br>", 
                            "Freq: ", format(freq, big.mark = ",")),
        marker = list(color = '#2C3E50')) %>% 
  layout(title = list(
    text = paste0(
      "<b>",
      document_cat,
      " - Word Frequency",
      "</b>",
      "<br>",
      "Top 50 Words"
    ),
    font = title_font_list
  ), 
    xaxis = list(title = 'Word Frequency'), # Ensuring terms are treated as categorical data
    yaxis = list(title = '', type = 'category'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "y unified"
  ) %>% 
  config(displaylogo = FALSE)
```

## TF-IDF

```{r}
# Assuming you have multiple documents in your corpus
tfidf <- weightTfIdf(tdm)
tfidf_matrix <- as.matrix(tfidf)
# Find the highest values by row
top_tfidf <- apply(tfidf_matrix, 1, max)

# Create a data frame for plotting
tfidf_df <- data.frame(term = names(top_tfidf), score = top_tfidf)
tfidf_df <- tfidf_df[order(tfidf_df$score, decreasing = TRUE), ]
```

```{r}
top_tf_idf <- tfidf_df %>% 
  dplyr::arrange(desc(score)) %>% 
  dplyr::slice_head(n = 50)

# order bars 
top_tf_idf$term <- factor(top_tf_idf$term,
                          levels = unique(top_tf_idf$term)[order(top_tf_idf$score, decreasing = F)])

# Define a custom color scale and reverse it
custom_scale <- list(c(0, "#ef476f"), c(0.5, "#ffd166"), c(1, "#06d6a0"))

plot_ly(data = top_tf_idf, y = ~term, x = ~score, 
        type = 'bar', 
        hoverinfo = "text", 
        textposition = "auto", 
        hovertext = ~paste0("<b>",term,"</b>","<br>
                            Top 50 Words", 
                            "Score: ", round(score, digits = 2)),
        marker = list(color = ~score, 
                      colorbar = list(title = "TF-IDF<br>Score"), 
                      colorscale = custom_scale)) %>% 
  layout(title = list(
    text = paste0(
      "<b>",
      document_cat,
      " - TF-IDF Scores",
      "</b>",
      "<br>",
      "Top 50 Words"
    ),
    font = title_font_list
  ), 
    xaxis = list(title = "TF-IDF Score"), # Ensuring terms are treated as categorical data
    yaxis = list(title = '', type = 'category'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "y unified"
  ) %>% 
  config(displaylogo = FALSE)
``` 


## Wordcloud 

```{r}
# Plot the word cloud
wordcloud(
  words = df$word,
  freq = df$freq,
  min.freq = 500, # increasing min.freq to exclude more less frequent words 
  scale = c(2, 0.5), #adjusting scale param to affect the size ratio between most/least frequent words, helps fit more words 
  
  colors = brewer.pal(8, "Dark2")
)

# wordcloud2(df,color = "random-light" )
```

## Text Complexity 

```{r}
corpus_text <- corpus(valid_operable)

readability_indexs <- c(
  "Flesch",
  "Flesch.Kincaid",
  "ARI",
  "Coleman",
  "Coleman.Liau.ECP",
  "Coleman.Liau.grade",
  "Coleman.Liau.short",
  "RIX",
  "SMOG",
  "meanSentenceLength",
  "meanWordSyllables",
  "FOG.NRI"
)

readability_results <- textstat_readability(
  corpus_text, 
  measure = readability_indexs)

# Convert to long format for plotting 
readability_results_plot <- readability_results %>% 
  tidyr::pivot_longer(cols = c(Flesch:SMOG, FOG.NRI), values_to = "Score", names_to = "Metric") %>% 
  dplyr::select(document, Metric, Score, meanSentenceLength, meanWordSyllables)

# Plot Scores 

plot_ly(data = readability_results_plot, x = ~Metric, y = ~Score, 
        type = 'box',
        color = I("#F39C12")) %>% 
  layout(
    title = list(
      text = paste0("<b>", document_cat, "</b><br>Readabilty Scores", "</b>"),
      font = title_font_list
    ),
    xaxis = list(title = 'Word Frequency', type = 'category'), # Ensuring terms are treated as categorical data
    yaxis = list(title = 'Score'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "x unified"
  ) %>% 
  config(displaylogo = FALSE)
```

## Topic Model 

```{r}
dfm_valid_operable <- tokens(corpus_text, remove_punct = TRUE, remove_numbers = TRUE) %>%
  tokens_remove(stopwords("en")) %>%
  dfm() %>%
  dfm_trim(min_termfreq = 4, max_docfreq = 10)

# fit LDA model 
LDA_fit_20 <- quanteda::convert(dfm_valid_operable, to = "topicmodels") |> 
  LDA(k = 20)

# Print Topic Models
get_terms(LDA_fit_20, 5)
```

# Ratified agreements (ACUT) {.tabset .tabset-fade .tabset-pills}

```{r}
document_cat <- "Ratified Agreements"
```

```{r}
# Valid and Operable (VCUT)
ratified <- tm::Corpus(
  x = tm::DirSource("./00_data/original_treaties_USA/justdocsACUT/"),
  readerControl = list(
    reader = readPlain,
    language = "en",
    load = FALSE
  )
)

ratified
```

```{r}
# First, preprocess the data: remove common stop words, punctuation, and convert to lowercase
corpus <- ratified
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeWords, stopwords("english"))

# Create a term-document matrix
tdm <- TermDocumentMatrix(corpus)

# Convert the matrix to a data frame suitable for plotting
m <- as.matrix(tdm)
word_freqs <- sort(rowSums(m), decreasing = TRUE)
df <- data.frame(word = names(word_freqs), freq = word_freqs)
```


## Text Statistics 

```{r}
top_words <- df %>% dplyr::arrange(desc(freq)) %>% dplyr::slice_head(n = 50)

# order bars 
top_words$word <- factor(top_words$word,
                                      levels = unique(top_words$word)[order(top_words$freq, decreasing = F)])

plot_ly(data = top_words, y = ~word, x = ~freq, 
        type = 'bar', 
        hoverinfo = "text", 
        textposition = "auto", 
        hovertext = ~paste0("<b>",word,"</b>","<br>", 
                            "Freq: ", format(freq, big.mark = ",")),
        marker = list(color = '#2C3E50')) %>% 
  layout(
    title = list(
      text = paste0("<b>", document_cat, " - Word Frequency", "</b>", "<br>", 
                    "Top 50 Words"),
      font = title_font_list
    ),
    xaxis = list(title = 'Word Frequency'), # Ensuring terms are treated as categorical data
    yaxis = list(title = '', type = 'category'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "y unified"
  ) %>% 
  config(displaylogo = FALSE)
```

## TF-IDF

```{r}
# Assuming you have multiple documents in your corpus
tfidf <- weightTfIdf(tdm)
tfidf_matrix <- as.matrix(tfidf)
# Find the highest values by row
top_tfidf <- apply(tfidf_matrix, 1, max)

# Create a data frame for plotting
tfidf_df <- data.frame(term = names(top_tfidf), score = top_tfidf)
tfidf_df <- tfidf_df[order(tfidf_df$score, decreasing = TRUE), ]
```

```{r}
top_tf_idf <- tfidf_df %>% 
  dplyr::arrange(desc(score)) %>% 
  dplyr::slice_head(n = 50)

# order bars 
top_tf_idf$term <- factor(top_tf_idf$term,
                          levels = unique(top_tf_idf$term)[order(top_tf_idf$score, decreasing = F)])

# Define a custom color scale and reverse it
custom_scale <- list(c(0, "#ef476f"), c(0.5, "#ffd166"), c(1, "#06d6a0"))

plot_ly(data = top_tf_idf, y = ~term, x = ~score, 
        type = 'bar', 
        hoverinfo = "text", 
        textposition = "auto", 
        hovertext = ~paste0("<b>",term,"</b>","<br>
                            Top 50 Words", 
                            "Score: ", round(score, digits = 2)),
        marker = list(color = ~score, 
                      colorbar = list(title = "TF-IDF<br>Score"), 
                      colorscale = custom_scale)) %>% 
  layout(title = list(
    text = paste0(
      "<b>",
      document_cat,
      " - TF-IDF Scores",
      "</b>",
      "<br>",
      "Top 50 Words"
    ),
    font = title_font_list
  ), 
    xaxis = list(title = "TF-IDF Score"), # Ensuring terms are treated as categorical data
    yaxis = list(title = '', type = 'category'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "y unified"
  ) %>% 
  config(displaylogo = FALSE)
``` 


## Wordcloud 

```{r}
# Plot the word cloud
wordcloud(
  words = df$word,
  freq = df$freq,
  min.freq = 200, # increasing min.freq to exclude more less frequent words 
  scale = c(2, 0.5), #adjusting scale param to affect the size ratio between most/least frequent words, helps fit more words 
  
  colors = brewer.pal(8, "Dark2")
)

# wordcloud2(df,color = "random-light" )
```

## Text Complexity 

```{r}
corpus_text <- corpus(valid_operable)

readability_indexs <- c(
  "Flesch",
  "Flesch.Kincaid",
  "ARI",
  "Coleman",
  "Coleman.Liau.ECP",
  "Coleman.Liau.grade",
  "Coleman.Liau.short",
  "RIX",
  "SMOG",
  "meanSentenceLength",
  "meanWordSyllables",
  "FOG.NRI"
)

readability_results <- textstat_readability(
  corpus_text, 
  measure = readability_indexs)

# Convert to long format for plotting 
readability_results_plot <- readability_results %>% 
  tidyr::pivot_longer(cols = c(Flesch:SMOG, FOG.NRI), values_to = "Score", names_to = "Metric") %>% 
  dplyr::select(document, Metric, Score, meanSentenceLength, meanWordSyllables)

# Plot Scores 

plot_ly(data = readability_results_plot, x = ~Metric, y = ~Score, 
        type = 'box',
        color = I("#F39C12")) %>% 
  layout(
    title = list(
      text = paste0("<b>", document_cat, "</b><br>Readabilty Scores", "</b>"),
      font = title_font_list
    ),
    xaxis = list(title = 'Word Frequency', type = 'category'), # Ensuring terms are treated as categorical data
    yaxis = list(title = 'Score'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "x unified"
  ) %>% 
  config(displaylogo = FALSE)
```

## Topic Model 

```{r}
dfm_ratified <- tokens(corpus_text, remove_punct = TRUE, remove_numbers = TRUE) %>%
  tokens_remove(stopwords("en")) %>%
  dfm() %>%
  dfm_trim(min_termfreq = 4, max_docfreq = 10)

# fit LDA model 
LDA_fit_20 <- quanteda::convert(dfm_ratified, to = "topicmodels") |> 
  LDA(k = 20)

# Print Topic Models
get_terms(LDA_fit_20, 5)
```

# Rejected (RCUT) {.tabset .tabset-fade .tabset-pills}

```{r}
document_cat <- "Rejected Agreements"
```

```{r}
# Valid and Operable (VCUT)
rejected <- tm::Corpus(
  x = tm::DirSource("./00_data/original_treaties_USA/justdocsRCUT/"),
  readerControl = list(
    reader = readPlain,
    language = "en",
    load = FALSE
  )
)

rejected
```

```{r}
# First, preprocess the data: remove common stop words, punctuation, and convert to lowercase
corpus <- rejected
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeWords, stopwords("english"))

# Create a term-document matrix
tdm <- TermDocumentMatrix(corpus)

# Convert the matrix to a data frame suitable for plotting
m <- as.matrix(tdm)
word_freqs <- sort(rowSums(m), decreasing = TRUE)
df <- data.frame(word = names(word_freqs), freq = word_freqs)
```


## Text Statistics 

```{r}
top_words <- df %>% dplyr::arrange(desc(freq)) %>% dplyr::slice_head(n = 50)

# order bars 
top_words$word <- factor(top_words$word,
                                      levels = unique(top_words$word)[order(top_words$freq, decreasing = F)])

plot_ly(data = top_words, y = ~word, x = ~freq, 
        type = 'bar', 
        hoverinfo = "text", 
        textposition = "auto", 
        hovertext = ~paste0("<b>",word,"</b>","<br>", 
                            "Freq: ", format(freq, big.mark = ",")),
        marker = list(color = '#2C3E50')) %>% 
  layout(
    title = list(
      text = paste0("<b>", document_cat, " - Word Frequency", "</b>", "<br>", 
                    "Top 50 Words"),
      font = title_font_list
    ),
    xaxis = list(title = 'Word Frequency'), # Ensuring terms are treated as categorical data
    yaxis = list(title = '', type = 'category'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "y unified"
  ) %>% 
  config(displaylogo = FALSE)
```

## TF-IDF

```{r}
# Assuming you have multiple documents in your corpus
tfidf <- weightTfIdf(tdm)
tfidf_matrix <- as.matrix(tfidf)
# Find the highest values by row
top_tfidf <- apply(tfidf_matrix, 1, max)

# Create a data frame for plotting
tfidf_df <- data.frame(term = names(top_tfidf), score = top_tfidf)
tfidf_df <- tfidf_df[order(tfidf_df$score, decreasing = TRUE), ]
```

```{r}
top_tf_idf <- tfidf_df %>% 
  dplyr::arrange(desc(score)) %>% 
  dplyr::slice_head(n = 50)

# order bars 
top_tf_idf$term <- factor(top_tf_idf$term,
                          levels = unique(top_tf_idf$term)[order(top_tf_idf$score, decreasing = F)])

# Define a custom color scale and reverse it
custom_scale <- list(c(0, "#ef476f"), c(0.5, "#ffd166"), c(1, "#06d6a0"))

plot_ly(data = top_tf_idf, y = ~term, x = ~score, 
        type = 'bar', 
        hoverinfo = "text", 
        textposition = "auto", 
        hovertext = ~paste0("<b>",term,"</b>","<br>
                            Top 50 Words", 
                            "Score: ", round(score, digits = 2)),
        marker = list(color = ~score, 
                      colorbar = list(title = "TF-IDF<br>Score"), 
                      colorscale = custom_scale)) %>% 
  layout(title = list(
    text = paste0(
      "<b>",
      document_cat,
      " - TF-IDF Scores",
      "</b>",
      "<br>",
      "Top 50 Words"
    ),
    font = title_font_list
  ), 
    xaxis = list(title = "TF-IDF Score"), # Ensuring terms are treated as categorical data
    yaxis = list(title = '', type = 'category'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "y unified"
  ) %>% 
  config(displaylogo = FALSE)
``` 


## Wordcloud 

```{r}
# Plot the word cloud
wordcloud(
  words = df$word,
  freq = df$freq,
  min.freq = 200, # increasing min.freq to exclude more less frequent words 
  scale = c(2, 0.5), #adjusting scale param to affect the size ratio between most/least frequent words, helps fit more words 
  
  colors = brewer.pal(8, "Dark2")
)

# wordcloud2(df,color = "random-light" )
```

## Text Complexity 

```{r}
corpus_text <- corpus(valid_operable)

readability_indexs <- c(
  "Flesch",
  "Flesch.Kincaid",
  "ARI",
  "Coleman",
  "Coleman.Liau.ECP",
  "Coleman.Liau.grade",
  "Coleman.Liau.short",
  "RIX",
  "SMOG",
  "meanSentenceLength",
  "meanWordSyllables",
  "FOG.NRI"
)

readability_results <- textstat_readability(
  corpus_text, 
  measure = readability_indexs)

# Convert to long format for plotting 
readability_results_plot <- readability_results %>% 
  tidyr::pivot_longer(cols = c(Flesch:SMOG, FOG.NRI), values_to = "Score", names_to = "Metric") %>% 
  dplyr::select(document, Metric, Score, meanSentenceLength, meanWordSyllables)

# Plot Scores 

plot_ly(data = readability_results_plot, x = ~Metric, y = ~Score, 
        type = 'box',
        color = I("#F39C12")) %>% 
  layout(
    title = list(
      text = paste0("<b>", document_cat, "</b><br>Readabilty Scores", "</b>"),
      font = title_font_list
    ),
    xaxis = list(title = 'Word Frequency', type = 'category'), # Ensuring terms are treated as categorical data
    yaxis = list(title = 'Score'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "x unified"
  ) %>% 
  config(displaylogo = FALSE)
```

## Topic Model 

```{r}
dfm_rejected <- tokens(corpus_text, remove_punct = TRUE, remove_numbers = TRUE) %>%
  tokens_remove(stopwords("en")) %>%
  dfm() %>%
  dfm_trim(min_termfreq = 4, max_docfreq = 10)

# fit LDA model 
LDA_fit_20 <- quanteda::convert(dfm_rejected, to = "topicmodels") |> 
  LDA(k = 20)

# Print Topic Models
get_terms(LDA_fit_20, 5)
```

# Unratified (UCUT) {.tabset .tabset-fade .tabset-pills}

```{r}
document_cat <- "Unratified Agreements"
```

```{r}
# Valid and Operable (VCUT)
unratified <- tm::Corpus(
  x = tm::DirSource("./00_data/original_treaties_USA/justdocsUCUT/"),
  readerControl = list(
    reader = readPlain,
    language = "en",
    load = FALSE
  )
)

unratified
```

```{r}
# First, preprocess the data: remove common stop words, punctuation, and convert to lowercase
corpus <- unratified
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeWords, stopwords("english"))

# Create a term-document matrix
tdm <- TermDocumentMatrix(corpus)

# Convert the matrix to a data frame suitable for plotting
m <- as.matrix(tdm)
word_freqs <- sort(rowSums(m), decreasing = TRUE)
df <- data.frame(word = names(word_freqs), freq = word_freqs)
```


## Text Statistics 

```{r}
top_words <- df %>% dplyr::arrange(desc(freq)) %>% dplyr::slice_head(n = 50)

# order bars 
top_words$word <- factor(top_words$word,
                                      levels = unique(top_words$word)[order(top_words$freq, decreasing = F)])

plot_ly(data = top_words, y = ~word, x = ~freq, 
        type = 'bar', 
        hoverinfo = "text", 
        textposition = "auto", 
        hovertext = ~paste0("<b>",word,"</b>","<br>", 
                            "Freq: ", format(freq, big.mark = ",")),
        marker = list(color = '#2C3E50')) %>% 
  layout(
    title = list(
      text = paste0("<b>", document_cat, " - Word Frequency", "</b>", "<br>", 
                    "Top 50 Words"),
      font = title_font_list
    ),
    xaxis = list(title = 'Word Frequency'), # Ensuring terms are treated as categorical data
    yaxis = list(title = '', type = 'category'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "y unified"
  ) %>% 
  config(displaylogo = FALSE)
```

## TF-IDF

```{r}
# Assuming you have multiple documents in your corpus
tfidf <- weightTfIdf(tdm)
tfidf_matrix <- as.matrix(tfidf)
# Find the highest values by row
top_tfidf <- apply(tfidf_matrix, 1, max)

# Create a data frame for plotting
tfidf_df <- data.frame(term = names(top_tfidf), score = top_tfidf)
tfidf_df <- tfidf_df[order(tfidf_df$score, decreasing = TRUE), ]
```

```{r}
top_tf_idf <- tfidf_df %>% 
  dplyr::arrange(desc(score)) %>% 
  dplyr::slice_head(n = 50)

# order bars 
top_tf_idf$term <- factor(top_tf_idf$term,
                          levels = unique(top_tf_idf$term)[order(top_tf_idf$score, decreasing = F)])

# Define a custom color scale and reverse it
custom_scale <- list(c(0, "#ef476f"), c(0.5, "#ffd166"), c(1, "#06d6a0"))

plot_ly(data = top_tf_idf, y = ~term, x = ~score, 
        type = 'bar', 
        hoverinfo = "text", 
        textposition = "auto", 
        hovertext = ~paste0("<b>",term,"</b>","<br>
                            Top 50 Words", 
                            "Score: ", round(score, digits = 2)),
        marker = list(color = ~score, 
                      colorbar = list(title = "TF-IDF<br>Score"), 
                      colorscale = custom_scale)) %>% 
  layout(title = list(
    text = paste0(
      "<b>",
      document_cat,
      " - TF-IDF Scores",
      "</b>",
      "<br>",
      "Top 50 Words"
    ),
    font = title_font_list
  ), 
    xaxis = list(title = "TF-IDF Score"), # Ensuring terms are treated as categorical data
    yaxis = list(title = '', type = 'category'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "y unified"
  ) %>% 
  config(displaylogo = FALSE)
``` 


## Wordcloud 

```{r}
# Plot the word cloud
wordcloud(
  words = df$word,
  freq = df$freq,
  min.freq = 200, # increasing min.freq to exclude more less frequent words 
  scale = c(2, 0.5), #adjusting scale param to affect the size ratio between most/least frequent words, helps fit more words 
  
  colors = brewer.pal(8, "Dark2")
)

# wordcloud2(df,color = "random-light" )
```

## Text Complexity 

```{r}
corpus_text <- corpus(valid_operable)

readability_indexs <- c(
  "Flesch",
  "Flesch.Kincaid",
  "ARI",
  "Coleman",
  "Coleman.Liau.ECP",
  "Coleman.Liau.grade",
  "Coleman.Liau.short",
  "RIX",
  "SMOG",
  "meanSentenceLength",
  "meanWordSyllables",
  "FOG.NRI"
)

readability_results <- textstat_readability(
  corpus_text, 
  measure = readability_indexs)

# Convert to long format for plotting 
readability_results_plot <- readability_results %>%
  tidyr::pivot_longer(
    cols = c(Flesch:SMOG, FOG.NRI),
    values_to = "Score",
    names_to = "Metric"
  ) %>%
  dplyr::select(document, Metric, Score, meanSentenceLength, meanWordSyllables)

# Plot Scores 

plot_ly(data = readability_results_plot, x = ~Metric, y = ~Score, 
        type = 'box',
        color = I("#F39C12")) %>% 
  layout(
    title = list(
      text = paste0("<b>", document_cat, "</b><br>Readabilty Scores", "</b>"),
      font = title_font_list
    ),
    xaxis = list(title = 'Word Frequency', type = 'category'), # Ensuring terms are treated as categorical data
    yaxis = list(title = 'Score'),
    font = font_list, 
    margin = margin_list,
    barmode = 'stack',
    hovermode = "x unified"
  ) %>% 
  config(displaylogo = FALSE)
```

## Topic Model 

```{r}
dfm_unratified <- tokens(corpus_text, remove_punct = TRUE, remove_numbers = TRUE) %>%
  tokens_remove(stopwords("en")) %>%
  dfm() %>%
  dfm_trim(min_termfreq = 4, max_docfreq = 10)

# fit LDA model 
LDA_fit_20 <- quanteda::convert(dfm_unratified, to = "topicmodels") |> 
  LDA(k = 20)

# Print Topic Models
get_terms(LDA_fit_20, 5)
```


